---
title: "ASM Time Series Project"
author: "Yifan Yao, Dinara Kurmangaliyeva"
date: "2025-01-02"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The final report summarizing all findings and analyses is available as `TimeSeries_FinalReport.pdf`.

Number of cars made in Spain. Monthly Data
```{r}
serie=window(ts(read.table("Turismos.dat")/1000,start=1994,freq=12))
serie
```


```{r}
plot(serie, main="Cars made in Spain",ylab="Thousands of Units")
abline(v=1994:2020,lty=3,col=4)
```


# 1. Identification:
a) Determine the needed transformations to make the series stationary. Justify the transformations
carried out using graphical and numerical results.
b) Analyze the ACF and PACF of the stationary series to identify at least two plausible models. Reason
about what features of the correlograms you use to identify these models.

## Tranforming series into a stationary one:
### Is Variance constant?
```{r}
boxplot(serie~floor(time(serie)))
```
```{r}
# matrix(serie, nr = 12)
# length(serie)/12  # 26

m<-apply(matrix(serie,nrow=12),2,mean)
v<-apply(matrix(serie,nrow=12),2,var)

plot(v~m) # there is an increase. so we have to take logrithm
abline(lm(v~m),col=2,lty=3)
```

It is not constant. There is an increase. so we have to take logrithm. 

```{r}
lnserie <- log(serie)
plot(lnserie, main="ln Cars made in Spain",ylab="Thousands of Units")
abline(v=1994:2020,lty=3,col=4)
```

Let’s check the same plots for the transformed series. 
```{r}
matrixlnserie <- matrix(lnserie,nrow=12)
boxplot(matrixlnserie)

lnm<-apply(matrix(matrixlnserie,nrow=12),2,mean) # ln m
lnv<-apply(matrix(matrixlnserie,nrow=12),2,var) # ln v

plot(lnv~lnm) # there is an increase. so we have to take logrithm
abline(lm(lnv~lnm),col=2,lty=3)
```
The first transformation is done.

### Is seasonality present?
Yes seasonality is present, and has a downfall every august.

```{r}
monthplot(lnserie) # August!!
ts.plot(matrix(lnserie,nrow=12),col=1:8)
```
```{r}
d12lnserie<-diff(lnserie,lag=12)

plot(d12lnserie)
abline(h=0)
abline(h=mean(d12lnserie),col=2)

monthplot(d12lnserie)

ts.plot(matrix(d12lnserie,nrow=12),col=1:8)
```


### Is the mean constant?

```{r}
# Mean seemingly constant equal to zero!!! 
# code, the same as before
plot(d12lnserie)
abline(h=0)
abline(h=mean(d12lnserie),col=2)
mean(d12lnserie) 
```
```{r}
var(lnserie) # 0.164901
var(d12lnserie) # 0.03583941
```


Check for over-differentiation: Take an extra differentiation and compare the variances.

```{r}
d1d12lnserie=diff(d12lnserie)
var(d1d12lnserie) # 0.04014573
```


Is the current series already stationary? Yes. 
```{r}
par(mar = c(5, 4, 4, 2))
acf(d12lnserie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,3)), main ="ACF(serie)")
pacf(d12lnserie,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,3)), main ="PACF(serie)")

# the ACF plot is also useful for identifying non-stationary time series. For a stationary time series, the ACF will drop to zero relatively quickly, while the ACF of non-stationary data decreases slowly.
```

## Final transformation for the Number of cars made in Spain series: $W_t=(1-B^{12})\log X_t$

# 2. Estimation:
a) Use R to estimate the identified models.
```{r}
par(mfrow=c(1,2))
acf(d12lnserie, ylim=c(-1,1), lag.max = 72,col=c(2,rep(1,11)),lwd=2)
pacf(d12lnserie, ylim=c(-1,1), lag.max = 72,col=c(rep(1,11),2),lwd=2)

```

## p: AR(3) 
## q: MA(6) 
## P: AR(2)
## Q: MA(1)
```{r}
(mod1=arima(lnserie,order=c(3,0,0),seasonal=list(order=c(0,1,1),period=12))) #ar(3) for regular part
(mod2=arima(lnserie,order=c(0,0,6),seasonal=list(order=c(0,1,1),period=12))) #ma(6) for refular part
(mod3=arima(lnserie,order=c(1,0,1),seasonal=list(order=c(0,1,1),period=12))) #simplest model arma(1,1) and seasonal part is the same
```

# 3. Validation:
```{r}
#################Validation#################################
validation=function(model){
  s=frequency(get(model$series))
  resi=model$residuals
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  #Residuals plot
  plot(resi,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resi),3*sd(resi)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resi)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resi)
  qqline(resi,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resi,breaks=20,freq=FALSE)
  curve(dnorm(x,mean=mean(resi),sd=sd(resi)),col=2,add=T)
  
  
  #ACF & PACF of residuals
  par(mfrow=c(1,2))
  acf(resi,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=2)
  pacf(resi,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=2)
  par(mfrow=c(1,1))
  
  #Ljung-Box p-values
  par(mar=c(2,2,1,1))
  tsdiag(model,gof.lag=7*s)
  cat("\n--------------------------------------------------------------------\n")
  print(model)
  
  #Stationary and Invertible
  cat("\nModul of AR Characteristic polynomial Roots: ", 
      Mod(polyroot(c(1,-model$model$phi))),"\n")
  cat("\nModul of MA Characteristic polynomial Roots: ",
      Mod(polyroot(c(1,model$model$theta))),"\n")
  
  suppressMessages(require(forecast,quietly=TRUE,warn.conflicts=FALSE))
  plot(model)
  
  #Model expressed as an MA infinity (psi-weights)
  psis=ARMAtoMA(ar=model$model$phi,ma=model$model$theta,lag.max=36)
  names(psis)=paste("psi",1:36)
  cat("\nPsi-weights (MA(inf))\n")
  cat("\n--------------------\n")
  print(psis[1:24])
  
  #Model expressed as an AR infinity (pi-weights)
  pis=-ARMAtoMA(ar=-model$model$theta,ma=-model$model$phi,lag.max=36)
  names(pis)=paste("pi",1:36)
  cat("\nPi-weights (AR(inf))\n")
  cat("\n--------------------\n")
  print(pis[1:24])
   
  cat("\nDescriptive Statistics for the Residuals\n")
  cat("\n----------------------------------------\n") 
  
  suppressMessages(require(fBasics,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  #print(basicStats(resi))
  
  ## Add here complementary tests (use with caution!)
  ##---------------------------------------------------------
  cat("\nNormality Tests\n")
  cat("\n--------------------\n")
 
  ##Shapiro-Wilks Normality test
  print(shapiro.test(resi))

  suppressMessages(require(nortest,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  print(ad.test(resi))
  
  suppressMessages(require(tseries,quietly=TRUE,warn.conflicts=FALSE))
  ##Jarque-Bera test
  print(jarque.bera.test(resi))
  
  cat("\nHomoscedasticity Test\n")
  cat("\n--------------------\n")
  suppressMessages(require(lmtest,quietly=TRUE,warn.conflicts=FALSE))
  ##Breusch-Pagan test
  obs=get(model$series)
  print(bptest(resi~I(obs-resi)))
  
  cat("\nIndependence Tests\n")
  cat("\n--------------------\n")
  
  ##Durbin-Watson test
  print(dwtest(resi~I(1:length(resi))))
  
  ##Ljung-Box test
  cat("\nLjung-Box test\n")
  print(t(apply(matrix(c(1:4,(1:4)*s)),1,function(el) {
    te=Box.test(resi,type="Ljung-Box",lag=el)
    c(lag=(te$parameter),statistic=te$statistic[[1]],p.value=te$p.value)})))
  
}

################# Fi Validation #################################
```

a) Perform the complete analysis of residuals, justifying all assumptions made. Use the corresponding
tests and graphical results.

b) Include analysis of the expressions of the AR and MA infinite models, discuss if they are causal
and/or invertible and report some adequacy measures.

```{r}
#Model1
validation(mod1)

```

```{r}
#Model2
validation(mod2)
```


```{r}
#Model3
validation(mod3)
```



## c) Check the stability of the proposed models and evaluate their capability of prediction, reserving the last 12 observations.
d) Select the best model for forecasting.

### model 1
```{r}
# model 1
ultim=c(2018,12)
pdq=c(3,0,0)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))
```
```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```
```{r}
obs=window(serie,start=ultim+c(0,1))
pr=window(pr,start=ultim+c(0,1))
ts(data.frame(LowLim=tl[-1],Predic=pr,UpperLim=tu[-1],Observ=obs,Error=obs-pr,PercentError=(obs-pr)/obs),start=ultim+c(0,1),freq=12)
mod.RMSE1=sqrt(sum((obs-pr)^2)/12)
mod.MAE1=sum(abs(obs-pr))/12
mod.RMSPE1=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE1=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE1,"MAE"=mod.MAE1,"RMSPE"=mod.RMSPE1,"MAPE"=mod.MAPE1)
mCI1=mean(tu-tl)

cat("\nMean Length CI: ",mCI1)
```
### Final Prediction(prediction for Model 1)
```{r}
pred=predict(modA,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(ultim[1]-2,ultim[1]+3),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
(previs2=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0)))
```
### model 2 
```{r}
ultim=c(2018,12)
pdq=c(0,0,6)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))

```
```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```
```{r}
obs=window(serie,start=ultim+c(0,1))
pr=window(pr,start=ultim+c(0,1))
ts(data.frame(LowLim=tl[-1],Predic=pr,UpperLim=tu[-1],Observ=obs,Error=obs-pr,PercentError=(obs-pr)/obs),start=ultim+c(0,1),freq=12)
mod.RMSE2=sqrt(sum((obs-pr)^2)/12)
mod.MAE2=sum(abs(obs-pr))/12
mod.RMSPE2=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE2=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE2,"MAE"=mod.MAE2,"RMSPE"=mod.RMSPE2,"MAPE"=mod.MAPE2)
mCI2=mean(tu-tl)

cat("\nMean Length CI: ",mCI2)
```




prediction for model 2(not the final prediction)
```{r}
# pred=predict(modA,n.ahead=12)
# pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
# se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

# tl1<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
# tu1<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
# pr1<-ts(exp(pr),start=ultim+c(1,0),freq=12)

# ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(ultim[1]-2,ultim[1]+3),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
# abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
# (previs3=window(cbind(tl1,pr1,tu1),start=ultim+c(1,0)))
```

### model 3
```{r}
ultim=c(2018,12)
pdq=c(1,0,1)
PDQ=c(0,1,1)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

(modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(modB=arima(lnserie2,order=pdq,seasonal=list(order=PDQ,period=12)))
```

```{r}
pred=predict(modB,n.ahead=12)
pr<-ts(c(tail(lnserie2,1),pred$pred),start=ultim,freq=12)
se<-ts(c(0,pred$se),start=ultim,freq=12)

#Intervals
tl<-ts(exp(pr-1.96*se),start=ultim,freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim,freq=12)
pr<-ts(exp(pr),start=ultim,freq=12)


ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```
```{r}
obs=window(serie,start=ultim+c(0,1))
pr=window(pr,start=ultim+c(0,1))
ts(data.frame(LowLim=tl[-1],Predic=pr,UpperLim=tu[-1],Observ=obs,Error=obs-pr,PercentError=(obs-pr)/obs),start=ultim+c(0,1),freq=12)
mod.RMSE3=sqrt(sum((obs-pr)^2)/12)
mod.MAE3=sum(abs(obs-pr))/12
mod.RMSPE3=sqrt(sum(((obs-pr)/obs)^2)/12)
mod.MAPE3=sum(abs(obs-pr)/obs)/12

data.frame("RMSE"=mod.RMSE1,"MAE"=mod.MAE3,"RMSPE"=mod.RMSPE3,"MAPE"=mod.MAPE3)
mCI3=mean(tu-tl)

cat("\nMean Length CI: ",mCI3)
```

```{r}
resul=data.frame(
  par=c(length(coef(mod1)),length(coef(mod2)), length(coef(mod3))),
  Sigma2Z=c(mod1$sigma2,mod2$sigma2, mod3$sigma2),
  AIC=c(AIC(mod1),AIC(mod2), AIC(mod3)),
  BIC=c(BIC(mod1),BIC(mod2), BIC(mod3)),
  RMSE=c(mod.RMSE2,mod.RMSE3, mod.RMSE1),
  MAE=c(mod.MAE1,mod.MAE2, mod.MAE3),
  RMSPE=c(mod.RMSPE1,mod.RMSPE2, mod.RMSPE3),
  MAPE=c(mod.MAPE1,mod.MAPE2, mod.MAPE3),
  meanLength=c(mCI1,mCI2, mCI3)
)

row.names(resul)=c("ARIMA(3,0,0)(0,0,1)12","ARIMA(0,0,6)(0,1,1)12", "ARMIA(1,0,1)(0,1,1)12")

resul
```

Based on the analysis above, we chose Model 1 and its final prediction is in the corresponding part above. 
